---
title: "Homework2"
author: "Toygar Emre"
date: "04 12 2020"
output: html_document
---


```{r}
#Upload libraries and read datas
suppressMessages(library(tidyverse))
suppressMessages(library(plot3D))
suppressMessages(library(rgl))

X.Train<-read.table(file="uWaveGestureLibrary_X_TRAIN") %>% as_tibble()
Y.Train<-read.table(file="uWaveGestureLibrary_Y_TRAIN") %>% as_tibble()
Z.Train<-read.table(file="uWaveGestureLibrary_Z_TRAIN") %>% as_tibble()
```

QUESTION PART A

```{r}
sort(unique(X.Train$V1)) #Find unique classes

```

Given class no, first row number(instance) for the class will be chosen as a sample.
```{r}
Random.Instance<-function(class.no){
  row.no<-first(which(X.Train$V1==class.no)) 
  Gesture<-as.data.frame(cbind(t(X.Train[row.no,]),t(Y.Train[row.no,]),t(Z.Train[row.no,]))) #For 3D purpose we bind columns
  colnames(Gesture)<-c("X","Y","Z")
  Gesture<-Gesture[-1,]#First row is class info. We do not need it
  rownames(Gesture)<-NULL
  Gesture.v<-as_tibble(apply(Gesture,2,cumsum))#Acceleration to velocity transformation
  
  return(Gesture.v)
}
par(mar=c(1,1,1,1))

```

1st Class

```{r}
res<-Random.Instance(1)
res %>% head
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z") 

#With below one new window will be opened and you can check 3D object with rotations.
#However for homework we will use the above one
#rgl::plot3d(x = res$X,y = res$Y,z = res$Z,type = "l" ,col = "blue",xlab = "x",ylab = "y",zlab = "z")

```

2nd Class
```{r}
res<-Random.Instance(2)
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z") 
```

3rd Class
```{r}
res<-Random.Instance(3)
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z") 

```

4th Class
```{r}
res<-Random.Instance(4)
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z") 
```

5th Class
```{r}
res<-Random.Instance(5)
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z") 
```

6th Class
```{r}
res<-Random.Instance(6)
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z") 
```

7th Class
```{r}
res<-Random.Instance(7)
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z") 
```

8th Class
```{r}
res<-Random.Instance(8)
plot3D::scatter3D(x = res$X,y = res$Y,z = res$Z,col = "blue",type="l",xlab = "x",ylab = "y",zlab = "z")
```

QUESTION PART B
```{r}
new.X.Train<-X.Train %>% rename(class=V1) %>% 
  mutate(time.series.id=row_number()) %>% 
  gather(key = "time.index",value = "X",-c(1,ncol(.)))
new.X.Train %>% head

new.Y.Train<-Y.Train %>% rename(class=V1) %>% 
  mutate(time.series.id=row_number()) %>% 
  gather(key = "time.index",value = "Y",-c(1,ncol(.)))

new.Z.Train<-Z.Train %>% rename(class=V1) %>% 
  mutate(time.series.id=row_number()) %>% 
  gather(key = "time.index",value = "Z",-c(1,ncol(.)))

Long.Data<-full_join(new.X.Train,new.Y.Train,by=c("class","time.series.id","time.index")) %>% 
  full_join(new.Z.Train,by=c("class","time.series.id","time.index")) %>% #Join 3 axis together
  separate(time.index,c("V","time.index"),sep = "(?<=V) ?(?=[0-9])") %>% select(-V) %>% #Column names are V2,V3... 
  mutate(time.index=as.numeric(time.index)-1) %>% #we remove V and stay with the number
  arrange(time.series.id,class,time.index) %>% 
  select(time.series.id,time.index,X,Y,Z,class)

Long.Data %>% head
Long.Data %>% tail

```

```{r}
#Apply PCA
pca.data<-Long.Data %>% select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.data,center = T,scale. = T)$rotation %>% as_tibble()
pca.res

```

```{r}
#Create PC1 values using PCA weights
PC1<-pca.data%*%pca.res$PC1 %>% as_tibble(.name_repair = "minimal")
colnames(PC1)<-"PC1"

pca.final<-Long.Data %>% select(time.series.id,time.index,class) %>% bind_cols(PC1) 
pca.final %>% head
```


```{r}
#Randomly select 2 time series from each class
random.sample<-2
pca.test<-pca.final %>% select(class,time.series.id) %>% unique() %>% 
  group_by(class) %>% 
  mutate(row.no=row_number(),
         maxrow.no=max(row.no)) %>% 
  filter(row.no %in% sample(1:maxrow.no[1],random.sample)) %>% 
  ungroup()

pca.final %>% filter(time.series.id %in% pca.test$time.series.id) %>% 
  mutate(class=as.character(class)) %>% 
  ggplot(aes(x=time.index, y=PC1,group=time.series.id, colour=class)) +
  geom_line()

```

This is a bit confusing so we remove time.series.id info. In this way, two samples from each class will create areas.

```{r}
pca.final %>% filter(time.series.id %in% pca.test$time.series.id) %>% 
  mutate(class=as.character(class)) %>% 
  ggplot(aes(x=time.index, y=PC1, colour=class)) +
  geom_line()

```


With geom_smooth it may be easy to visualize

```{r}
pca.final %>% filter(time.series.id %in% pca.test$time.series.id) %>% 
  mutate(class=as.character(class)) %>% 
  ggplot(aes(x=time.index, y=PC1, colour=class)) +
  geom_smooth(stat = "identity")

```

We think classes can be separated with PCA as can be seen from the plots. It looks like each of them has unique pattern.


QUESTION PART C

1st Class
```{r}
pca.tmp<-Long.Data %>% filter(class==1) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)
```

2nd Class
```{r}
pca.tmp<-Long.Data %>% filter(class==2) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)
```

3rd Class
```{r}
pca.tmp<-Long.Data %>% filter(class==3) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)
```


4th Class
```{r}
pca.tmp<-Long.Data %>% filter(class==4) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)
```

5th Class
```{r}
pca.tmp<-Long.Data %>% filter(class==5) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)
```

6th Class
```{r}
pca.tmp<-Long.Data %>% filter(class==6) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)

```

7th Class
```{r}
pca.tmp<-Long.Data %>% filter(class==7) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)
```

8th Class
```{r}
pca.tmp<-Long.Data %>% filter(class==8) %>% 
  select(X,Y,Z) %>% as.matrix()
pca.res<-prcomp(pca.tmp,center = T,scale.=T)
pca.res$rotation
summary(pca.res)
```

With first component, more than 50% of variance is covered for each class.
I think this is an important indication that pca is useful to if one wants to reduce feature space for this type of gesture data.
There are some patterns in the values of first components.
For example, first components of 3rd and 4th classes; 2nd, 5th and 7th classes; 1st and 8th classes are similar to each other.


QUESTION PART D

```{r}
X.Train.dist<-as.matrix(X.Train[,-1])
Y.Train.dist<-as.matrix(Y.Train[,-1])
Z.Train.dist<-as.matrix(Z.Train[,-1])
classes<-X.Train[,1]
colnames(classes)<-"class"

colnames(X.Train.dist)<-paste0("V",1:ncol(X.Train.dist))
colnames(Y.Train.dist)<-paste0("V",1:ncol(Y.Train.dist))
colnames(Z.Train.dist)<-paste0("V",1:ncol(Z.Train.dist))

#Calculate distances between time series (rows)
dist.X<-dist(x=X.Train.dist,method = "euclidean",diag = T,upper = T) %>% as.matrix()
dist.Y<-dist(x=Y.Train.dist,method = "euclidean",diag = T,upper = T) %>% as.matrix()
dist.Z<-dist(x=Z.Train.dist,method = "euclidean",diag = T,upper = T) %>% as.matrix()

dist.all<-dist.X+dist.Y+dist.Z #Sum them up

#Multidimensional scaling to represent on 2d space
Scaled2d<-cmdscale(dist.all,k = 2) %>% as_tibble(.name_repair="minimal")
colnames(Scaled2d)<-c("X","Y")
Scaled2d %>% head()
```

Visualization
```{r}
Scaled2d %>%
  bind_cols(classes) %>% 
  mutate(class=as.character(class)) %>% 
  ggplot(aes(x=X,y=Y,colour=class))+
  geom_point(size=3)

```

As can be seen from the plot, classes are grouped together meaning that samples within classes are close.
In other words, multidimensional scaling works for separating classes.
